BlockTensorDecompositions.jl: A Unified Constrained Tensor Decomposition Julia Package

### 2025-01-14 paper/paper.qmd
continue on Computing Gradients section

---
title: "BlockTensorDecompositions.jl: A Unified Constrained Tensor Decomposition Julia Package"
subtitle: ""
authors:
  - name: "Nicholas J. E. Richardson"
    affiliations:
      - ref: math
  - name: "Noah Marusenko"
    affiliations:
      - ref: cs
  - name: "Michael P. Friedlander"
    affiliations:
      - ref: mathcs
affiliations:
  - id: math
    name: "Department of Mathematics"
  - id: cs
    name: "Department of Computer Science"
  - id: mathcs
    name: "Departments of Mathematics and Computer Science\n"
date: ""
date-format: "D MMMM YYYY"
slide-number: true
categories: [tensors, signal processing, julia]
execute:
  freeze: true
bibliography: references.bib
csl: citationstyles/ieee-compressed-in-text-citations.csl # Citation Style
callout-appearance: minimal
callout-icon: false
format:
  typst:
    toc: true
    section-numbering: 1.1.a.i
    keep-typ: true
    mainfont: "Libertinus Serif"
editor:
  render-on-save: true
#appendix-style: plain
# nocite: |
#   @firstCitation
#embed-resources: true ## uncomment when sharing HTML file
---

@@

@@

The $n$-slice dot product $\cdot_n$ between two tensors $A\in\mathbb{R}^{I_{1},\dots, I_{n-1},J,I_{n+1},\dots, I_{N}}$ and $B\in\mathbb{R}^{I_{1},\dots, I_{n-1},K,I_{n+1},\dots, I_{N}}$ returns a matrix $(A \cdot_n B)\in\mathbb{R}^{J\times K}$ with entries

$$
(A \cdot_n B)[j, k] = {\sum_{i_{1}\dots i_{n-1}i_{n+1}\dots i_{N}}} A[i_{1},\dots, i_{n-1},j,i_{n+1},\dots, i_{N}]B[i_{1},\dots, i_{n-1},k,i_{n+1},\dots, i_{N}].
$$

This product can also be thought of as taking the dot product between all pairs of $n$th order slices of $A$ and $B$: $(A \cdot_n B)[j, k] = A_j \cdot B_k$.

@@

    # Initialize the output tensor
    C = zeros(size(B)[1], sizeA[2:end]...)
    Cmat = reshape(C, size(B)[1], prod(sizeA[2:end]))

@@

@@

A tensor decomposition is a factorization of a tensor into multiple (usually smaller) tensors, that can be recombined into the original tensor. Computationally, we can think of a generic decomposition as storing factors $(A,B,C,...)$ and operations $(\times_a,\times_b,...)$ for combining them. This is what we do in BlockTensorDecomposition.jl.

```julia
struct GenericDecomposition{T, N} <: AbstractArray{T, N}
    factors::Tuple{Vararg{AbstractArray{T}}} # e.g. (A, B, C)
    contractions::Tuple{Vararg{Function}} # e.g. (×₁, ×₂)
end
# Y = A ×₁ B ×₂ C
array(G::GenericDecomposition) = multifoldl(contractions(G), factors(G))
```

@@

```julia
function multifoldl(ops, args)
    @assert (length(ops) + 1) == length(args)
    x = args[begin]
    for (op, arg) in zip(ops, args[begin+1:end])
        x = op(x, arg)
    end
    return x
end
```

@@

There are implemented in BlockTensorDecomposition.jl and can be called, for a third order tensor, with `Tucker((B, A₁, A₂, A₃))`, `Tucker1((B, A₁))`, and `CPDecomposition((A₁, A₂, A₃))`. These Julia `structs` store the tensor in its factored form. If the recombined tensor or particular entries are requested, Julia dispatches on the type of decomposition and calls a particular method of `array` or `getindex`. The implementations for efficient array construction and index access are provided below.

@@

function getindex(T::Tucker1, I::Vararg{Int})
    B, A = factors(T)
    i = I[1]
    J = I[begin+1:end] # J = (I₂, I₃, ..., I_N)
    return (@view A[i, :]) ⋅ view(B, :, J...)
end
```

@@

$$
\min_{A_1, \dots, A_N} \left\lVert g(A_1,\dots,A_N) - Y\right\rVert _F^2 \quad \text{s.t.} \quad (A_1,\dots,A_N) \in\mathcal{C}_1 \times \dots \times \mathcal{C}_N
$$ {#eq-constrained-least-squares}

@@

@@

Let $f(A_1, \dots, A_N) := \left\lVert g(A_1,\dots,A_N) - Y\right\rVert _F^2$ be the objective function we wish to minimize in @eq-constrained-least-squares. Following Xu and Yin [@xu_BlockCoordinateDescent_2013], the general approach we take to minimize $f$ is to apply block coordinate descent using each factor as a different block. Let $A_n^t$ be the $t$th iteration of the $n$th factor, and let

$$
f_n^t(A_n) := \left\lVert g(A_{1}^{t+1}, \dots, A_{n-1}^{t+1}, A_n, A_{n+1}^t, \dots , A_{N}^t)- Y \right\rVert _F^2
$$

@@

To ensure the code stays flexible, the main algorithm of BlockTensorDecomposition.jl, `factorize` is defined at a very high level.

<!--
```julia
factorize(Y; kwargs...) =
    _factorize(Y; (default_kwargs(Y; kwargs...))...)

"""
Inner level function once keyword arguments are set
"""
function _factorize(Y; kwargs...)
    decomposition, kwargs = initialize_decomposition(Y; kwargs...)
    previous, updateprevious! = initialize_previous(decomposition, Y; kwargs...)
    parameters, updateparameters! =
        initialize_parameters(decomposition, Y, previous; kwargs...)

    # more kwargs get added here, so need to return kwargs
    update!, kwargs = make_update!(decomposition, Y; kwargs...)

    stats_data, getstats =
        initialize_stats(decomposition, Y, previous, parameters; kwargs...)

    converged = make_converged(; kwargs...)

    # freeze the kwargs from a Dictionary to a NamedTuple for type stability
    kwargs = NamedTuple(kwargs)

    while !converged(stats_data; kwargs...)
        # Update the decomposition
        # This is one cycle of updates on each factor in the decomposition
        update!(decomposition; parameters...)

        # Update parameters used for the next update of the decomposition
        # this could be the next stepsize or other info used by update!
        updateparameters!(parameters, decomposition, previous)

        # Save stats
        push!(stats_data,
            getstats(decomposition, Y, previous, parameters, stats_data))

        # Update the one or two previous iterates
        # We do not save every iterate since that is too much memory
        updateprevious!(previous, parameters, decomposition)
    end

    # Perform one final constrain if requested
    if kwargs[:constrain_output]
        kwargs = finalconstrain!(decomposition; kwargs...)
        updateparameters!(parameters, decomposition, previous)
        push!(stats_data,
            getstats(decomposition, Y, previous, parameters, stats_data))
    end

Abstracting further, we can see the general structure. -->

```julia
function factorize(Y; kwargs...)
    decomposition, converged, update!, updateprevious!, updateparameters!, stats_data, getstats, kwargs = initialize(Y; kwargs...)

    while !converged(stats_data; kwargs...)
        update!(decomposition; parameters...)

        updateparameters!(parameters, decomposition, previous)

@@

        updateprevious!(previous, parameters, decomposition)
    end

    postprocess!(decomposition, parameters, stats_data, Y; kwargs...)

@@

@@

>>>>

---
title: "BlockTensorDecompositions.jl: A Unified Constrained Tensor Decomposition Julia Package"
subtitle: ""
authors:
  - name: "Nicholas J. E. Richardson"
    affiliations:
      - ref: math
  - name: "Noah Marusenko"
    affiliations:
      - ref: cs
  - name: "Michael P. Friedlander"
    affiliations:
      - ref: mathcs
affiliations:
  - id: math
    name: "Department of Mathematics"
  - id: cs
    name: "Department of Computer Science"
  - id: mathcs
    name: "Departments of Mathematics and Computer Science\n"
date: ""
date-format: "D MMMM YYYY"
slide-number: true
categories: [tensors, signal processing, julia]
execute:
  freeze: true
bibliography: references.bib
csl: citationstyles/ieee-compressed-in-text-citations.csl # Citation Style
callout-appearance: minimal
callout-icon: false
format:
  typst:
    toc: true
    section-numbering: 1.1.a.i
    keep-typ: true
    mainfont: "Libertinus Serif"
editor:
  render-on-save: true
#engine: julia
#appendix-style: plain
# nocite: |
#   @firstCitation
#embed-resources: true ## uncomment when sharing HTML file
---

@@

The transpose $A^{\top}\in\mathbb{R}^{J\times I}$ of a matrix $A\in\mathbb{R}^{I\times J}$ flips entries along the main diagonal: $A^{\top}[j,i] = A[i,j]$. In Julia, the transpose of a matrix is typed with a single apostrophe `A'`.

@@

The $n$-slice dot product $\cdot_n$ between two tensors $A\in\mathbb{R}^{K_{1},\dots, K_{n-1},I,K_{n+1},\dots, K_{N}}$ and $B\in\mathbb{R}^{K_{1},\dots, K_{n-1},J,K_{n+1},\dots, K_{N}}$ returns a matrix $(A \cdot_n B)\in\mathbb{R}^{I\times J}$ with entries

$$
(A \cdot_n B)[i, j] = {\sum_{k_{1}\dots k_{n-1}k_{n+1}\dots k_{N}}} A[k_{1},\dots, k_{n-1},i,k_{n+1},\dots, k_{N}]B[k_{1},\dots, k_{n-1},j,k_{n+1},\dots, k_{N}].
$$

This product can also be thought of as taking the dot product $(A \cdot_n B)[i, j] = A_i \cdot B_j$ between all pairs of $n$th order slices of $A$ and $B$, which exactly how BlockTensorDecomposition.jl defines the operation.

```julia
function slicewise_dot(A::AbstractArray, B::AbstractArray; dims=1)
    C = zeros(size(A, dims), size(B, dims))
    if A === B # use faster routine if they are the same
        return _slicewise_self_dot!(C, A; dims)
    end

    for (i, A_slice) in enumerate(eachslice(A; dims))
        for (j, B_slice) in enumerate(eachslice(B; dims))
            C[i, j] = A_slice ⋅ B_slice
        end
    end
    return C
end

function _slicewise_self_dot!(C, A; dims=1)
    enumerated_A_slices = enumerate(eachslice(A; dims))
    for (i, Ai_slice) in enumerated_A_slices
        for (j, Aj_slice) in enumerated_A_slices
            if i > j
                continue
            else # only compute the upper triangle entries of C
                C[i, j] = Ai_slice ⋅ Aj_slice
            end
        end
    end
    return Symmetric(C) # indexing C[2,1] points to the entry in C[1,2]
end
```

@@

    # Initialize the output tensor
    C = zeros(size(B, 1), sizeA[2:end]...)
    Cmat = reshape(C, size(B, 1), prod(sizeA[2:end]))

@@

::: {.callout-note}

If we were only working with a fixed order of tensors, we could have defined `×₁` entry-wise with `Tullio.jl`. The function definition `tullio×₁` below gives an example for order three tensors.

```julia
function tullio×₁(A::AbstractArray{_,3}, B::AbstractMatrix)
  @tullio C[i, j, k] := A[r, j, k] * B[i, r]
  return C
end
```

But we would need a new definition for each ordered tensor, or use Julia's meta programming to write a method for each order at runtime.

@@

A tensor decomposition is a factorization of a tensor into multiple (usually smaller) tensors, that can be recombined into the original tensor. To make a common interface for decompositions, we make an abstract subtype of Julia's `AbstractArray`, and subtype `AbstractDecomposition` for our concrete tensor decompositions.

```julia
abstract type AbstractDecomposition{T, N} <: AbstractArray{T, N} end
```

Computationally, we can think of a generic decomposition as storing factors $(A,B,C,...)$ and operations $(\times_a,\times_b,...)$ for combining them. This is what we do in BlockTensorDecomposition.jl.

```julia
struct GenericDecomposition{T, N} <: AbstractDecomposition{T, N}
    factors::Tuple{Vararg{AbstractArray{T}}} # e.g. (A, B, C)
    contractions::Tuple{Vararg{Function}} # e.g. (×₁, ×₂)
end
# Y = A ×₁ B ×₂ C
array(G::GenericDecomposition) = multifoldl(contractions(G), factors(G))
```

@@

```julia
function multifoldl(ops, args)
    @assert (length(ops) + 1) == length(args)
    x, xs... = args
    for (op, arg) in zip(ops, xs)
        x = op(x, arg)
    end
    return x
end
```

@@

There are implemented in BlockTensorDecomposition.jl and can be called, for a third order tensor, with `Tucker((B, A₁, A₂, A₃))`, `Tucker1((B, A₁))`, and `CPDecomposition((A₁, A₂, A₃))`. These Julia `structs` store the tensor in its factored form. We could define the contractions for these types and use the common interface provided by `array`, but it turns out we can reconstruct the whole tensor more efficiently. If the recombined tensor or particular entries are requested, Julia dispatches on the type of decomposition and calls a particular method of `array` or `getindex`. The implementations for efficient array construction and index access are provided below.

@@

function getindex(T::Tucker1, I::Vararg{Int})
    B, A = factors(T)
    i, J... = I # (i, J) = (I[1], I[begin+1:end])
    return (@view A[i, :]) ⋅ view(B, :, J...)
end
```

@@

$$
\min_{A_1, \dots, A_N} \frac{1}{2} \left\lVert g(A_1,\dots,A_N) - Y\right\rVert _F^2 \quad \text{s.t.} \quad (A_1,\dots,A_N) \in\mathcal{C}_1 \times \dots \times \mathcal{C}_N
$$ {#eq-constrained-least-squares}

@@

Note the problem would have the same solutions as simply using the objective $\left\lVert g(A_1,\dots,A_N) - Y\right\rVert$ without squaring and dividing by $2$. We define the objective in @eq-constrained-least-squares to make computing the function value and gradients faster.

@@

Let $f(A_1, \dots, A_N) := \frac{1}{2}\left\lVert g(A_1,\dots,A_N) - Y\right\rVert _F^2$ be the objective function we wish to minimize in @eq-constrained-least-squares. Following Xu and Yin [@xu_BlockCoordinateDescent_2013], the general approach we take to minimize $f$ is to apply block coordinate descent using each factor as a different block. Let $A_n^t$ be the $t$th iteration of the $n$th factor, and let

$$
f_n^t(A_n) := \frac{1}{2}\left\lVert g(A_{1}^{t+1}, \dots, A_{n-1}^{t+1}, A_n, A_{n+1}^t, \dots , A_{N}^t)- Y \right\rVert _F^2
$$

@@

To ensure the code stays flexible, the main algorithm of BlockTensorDecomposition.jl, `factorize`, is defined at a very high level.

```julia
factorize(Y; kwargs...) =
	_factorize(Y; (default_kwargs(Y; kwargs...))...)

"""
Inner level function once keyword arguments are set
"""
function _factorize(Y; kwargs...)
    decomposition, previous, updateprevious!, parameters, updateparameters!,
	update!, stats_data, getstats, converged, kwargs = initialize(Y, kwargs)

    while !converged(stats_data; kwargs...)
		# Usually one cycle of updates through each factor in the decomposition
        update!(decomposition; parameters...)

		# This could be the next stepsize or other info used by update!
        updateparameters!(parameters, decomposition, previous)

@@

		# Update one or two previous iterates. For example, used for momentum
        updateprevious!(previous, parameters, decomposition)
    end

    kwargs = postprocess!(decomposition, Y, previous, parameters, stats_data, updateparameters!, getstats, kwargs)

@@

Generally, we can use automatic differentiation on $f$ to compute gradients. Some care needs to be taken otherwise the forward or backwards pass will have to be recompiled every iteration since the factors are updated every iteration.

But for Tucker decompositions, we can compute gradients faster than what an automatic differentiation scheme would give, by taking advantage of symmetry and other computational shortcuts.

Starting with the Tucker-1 decomposition (@def-tucker-1-decomposition), we would like to compute $\nabla_B f(B, A)$ and $\nabla_A f(B, A)$ for $f(B, A) = \frac{1}{2} \left\lVert AB - Y\right\rVert_F^2$ for a given input $Y$. We have the gradient

$$
\nabla_B f(B, A) = A^{\top}(AB - Y) = (B \times_{1} A - Y) \times_1 A^\top
$$ {#eq-tucker-1-gradient-1}

by chain rule, but it is more efficient to calculate the gradient as

$$
\nabla_B f(B, A) = (A^{\top} A)B - A^{\top}Y = B \times_1 (A^{\top} A) - Y\times_1 A^{\top}.
$$ {#eq-tucker-1-gradient-2}

^[Seeing @eq-tucker-1-gradient-1 and @eq-tucker-1-gradient-2 written using the $1$-mode product shows how it is "backwards" to normal matrix-matrix multiplication.]For $A\in\mathbb{R}^{I\times R}$, $B\in\mathbb{R}^{R\times J\times K}$, and $Y\in\mathbb{R}^{I\times J \times K}$, @eq-tucker-1-gradient-1 requires

$$
\underbrace{ 2IJKR }_{ AB-Y }+\underbrace{ IJK(2I-1) }_{ A^\top (AB-Y) } \sim 2IJKR+2I^2JK
$$
floating point operations (FLOPS) whereas @eq-tucker-1-gradient-2 only uses

$$
\underbrace{ \frac{R(R+1)}{2}(2I-1) }_{ A^\top A } + \underbrace{ RJK(2I-1) }_{ A^\top Y } + \underbrace{ 2R^2JK }_{ (A^{\top} A)B - (A^{\top}Y) } \sim 2IJKR+2R^2JK + IR^2
$$

FLOPS^[Note we have the smaller factor ${R(R+1)}/{2}$ and not the expected $R^2$ number of entries needed to compute $A^\top A$. The product is a symmetric matrix so only the upper or lower triangle of entries needs to be computed.]. So for small ranks $R \ll I$, @eq-tucker-1-gradient-2 is cheaper.

A similar story can be said about $\nabla_A f(B, A)$ which is most efficiently computed as

$$
\nabla_A f(B, A) = A(B\cdot _1 B) - Y\cdot_1 B.
$$

The associated implementation with BlockTensorDecomposition.jl is shown below. We define a `make_gradient` which takes the decomposition, factor index `n`, and data tensor `Y`, and creates a function that computes the gradient for the same type of decomposition. This lets us manipulate the function that computes the gradient, rather than just the computed gradient.

```julia
function make_gradient(T::Tucker1, n::Integer, Y::AbstractArray; objective::L2, kwargs...)
    if n==0 # the core is the zeroth factor
        function gradient0(T::Tucker1; kwargs...)
            (B, A) = factors(T)
            AA = A'A
            YA = Y×₁A'
            grad = B×₁AA - YA
            return grad
        end
        return gradient0
    elseif n==1 # the matrix is the first factor
        function gradient1(T::Tucker1; kwargs...)
            (B, A) = factors(T)
            BB = slicewise_dot(B, B)
            YB = slicewise_dot(Y, B)
            grad = A*BB - YB
            return grad
        end
        return gradient1
    else
        error("No $(n)th factor in Tucker1")
    end
end
```

Similarly, we also have special methods for the Tucker and CP Decomposition.

The gradient with respect to the core for a full Tucker factorization is

$$
\nabla_B f(B, A_1, \dots, A_N) = B \bigtimes_n A_n^\top A_n - Y \bigtimes_{n} A_n^\top,
$$

and the gradient with respect to the matrix factor $A_m$ is

$$
\nabla_{A_m} f(B, A_1, \dots, A_N) = A_m (B \bigtimes_{n\neq m} A_n) \cdot_m (B \bigtimes_{n\neq m} A_n) - Y \cdot_m (B \bigtimes_{n\neq m} A_n).
$$

```julia
function make_gradient(T::Tucker, n::Integer, Y::AbstractArray; objective::L2, kwargs...)
    N = ndims(T)
    if n==0 # the core is the zeroth factor
        function gradient_core(T::AbstractTucker; kwargs...)
            C = core(T)
            matrices = matrix_factors(T)
            gram_matrices = map(A -> A'A, matrices) # gram matrices AA = A'A, BB = B'B...
            YAB = tuckerproduct(Y, adjoint.(matrices)) # Y ×₁ A' ×₂ B' ...
            grad = tuckerproduct(C, gram_matrices) - YAB
            return grad
        end
        return gradient_core

    elseif n in 1:N # the matrix factors start at m=1
        function gradient_matrix(T::AbstractTucker; kwargs...)
            matrices = matrix_factors(T)
            TExcludeAn = tuckerproduct(core(T), matrices; exclude=n)
            An = factor(T, n)
            grad = An*slicewise_dot(TExcludeAn, TExcludeAn; dims=n) - slicewise_dot(Y, TExcludeAn; dims=n)
            return grad
        end
        return gradient_matrix

    else
        error("No $(n)th factor in Tucker")
    end
end
```

```julia
function make_gradient(T::CPDecomposition, n::Integer, Y::AbstractArray; objective::L2, kwargs...)
    N = ndims(T)
    if n in 1:N # the matrix factors start at m=1
        function gradient_matrix(T::AbstractTucker; kwargs...)
            matrices = matrix_factors(T)
            TExcludeAn = tuckerproduct(core(T), matrices; exclude=n)
            An = factor(T, n)
            grad = An*slicewise_dot(TExcludeAn, TExcludeAn; dims=n) - slicewise_dot(Y, TExcludeAn; dims=n)
            return grad
        end
        return gradient_matrix

    else
        error("No $(n)th factor in Tucker")
    end
end
```

The function `tuckerproduct(B, (A₁, ..., Aₙ))` computes
$$
B \bigtimes_{n} A_n = \llbracket B; A_1,\dots,A_N \rrbracket,
$$

and can optionally "exclude" one of the matrix factors `tuckerproduct(B, (A₁, ..., Aₙ); exclude=m)` to compute

$$
B \bigtimes_{n\neq m} A_n = \llbracket B; A_1,\dots,A_{n-1}, \text{id}_{I_n}, A_{n+1},\dots,A_{N} \rrbracket
$$
where $B\in\mathbb{R}^{I_1\times \dots \times I_N}$.

@@

Similar to automatic differentiation, there exist "automatic Lipschitz" calculations to upper bound the Lipschitz constant of a function [@virmaux_lipschitz_2018].
